\clearpage
    \section{Getting started: immediate mode}
This section will describe the basic concepts behind the language,
such as reverse Polish notation and the stack, and introduce
using the language in immediate mode.   

\subsection{Immediate mode}
In immediate mode, each line of Angort is compiled and run straight
away. This is how the Angort executable starts when it is not
given an Angort script's filename on the command line. Angort will also
drop back to this mode if a script completes without \texttt{quit} 
being called.
Running the interpreter without a command line will give a prompt:
\begin{v}
1|0>
\end{v}
The two numbers are the number of garbage-collectable objects in the
system and the number of items on the stack, respectively.
The interpreter is in ``immediate mode'', as opposed to ``compilation
mode'' --- any text entered will be compiled to bytecode and run when
enter is pressed,
rather than being added to a function definition.

\subsection{Reverse Polish notation and the stack}
\index{stack}
Angort is a stack-based language:
there is a single stack containing
values, and most functions change the contents
of the stack in some way. For example,
\begin{v}
3
\end{v}
by itself will just put the value 3 on the stack. Then
\begin{v}
.
\end{v}
will pop the value from the top of the stack and print it.
\begin{v}
3 4 + .
\end{v}
will push 3 and 4 onto the stack, then add them together replacing them
with 7, and then print the 7. This kind of notation is often
referred to as \textbf{reverse Polish notation} (RPN) as opposed to the more common
\textbf{infix} notation. More complex expressions are built 
out of sequences of operations on the stack. For example, the expression
\[
\sin(5+32+\sqrt{43 \times 12})
\]
would be written as
\begin{v}
43 12 * sqrt 32 + 5 + sin
\end{v}
This is a little difficult at first, but rapidly becomes second
nature\footnote{Most old calculators work using RPN, and quite a few
of the more powerful programmables still do.}

\subsection{Running scripts}
Scripts are traditionally\footnote{A somewhat unusual word to use when
there is currently only one user of the language.} given the extension
\texttt{.ang}, but this appears nowhere in the codebase. To run a script,
create a file of Angort function definitions and commands and pass it as
an argument to the \texttt{angort} command. Each line will be parsed
and run in order, unless it is a function definition in which case the
compilation will occur at the end of the definition.

Note that each line of Angort (outside a function) is parsed and run
immediately. This means that multiline flow control constructs (such as loops
and \texttt{if..then}) will not work unless they are inside a function.
This is easily solved by using anonymous functions, as described in Section~\ref{multilineflow}.

At the end of a script, control will drop into immediate mode. This is often useful
when Angort is used to provide a domain-specific control language: the script sets up
a large number of commands and perhaps initialises native C++ libraries, and then
waits for the user's commands. If this is not required, terminate the script with
the \texttt{quit} command. This will cause Angort to exit with the return code 0.

\subsection{Words and functions}
\index{word}
Most of the Angort language consists of words which each perform
a single, isolated task -- there is very little syntax. Even common
``syntactic'' keywords like \texttt{if} (for conditions), \texttt{[]} 
(for creating lists) and \texttt{+} (for addition) operate primarily through
stack operations, although \texttt{if} does some more complex things at compile time
(such as calculating jump offsets). Code in Angort largely consists of many of these words
chained together to perform a task. These words can be combined into user-defined
functions.

In general, I will use ``word'' to refer to a single Angort built-in token,
particularily if it compiles to a single opcode; while ``function'' will refer
to user-defined functions. However, I may vary -- there is no real distinction, since
an Angort function call is itself a single opcode.

\subsection{Types and literals}
Angort is a dynamically typed language, with type coercion (weak typing).
The following types are available:
\indw{range}\indw{frange}\indw{[}\indw{]}
\begin{center}
\begin{tabular}{|l|p{2.7in}|p{2in}|}\hline
\textbf{Name} & \textbf{Definition} & \textbf{Example of literal} \\ \hline
None & The nil object, specifying ``no value'' & \texttt{none} \\
Integer & 32 or 64-bit integers depending on architecture & \texttt{5045} \\
Long & depends on architecture & \texttt{5045l,32L,0ffx, 45ao, 10111b, 3200ffffhl} \\
Float & 32-bit floats & \texttt{54.0} \\
Double & 64-bit floats & \texttt{54.0l,32.2L} \\
String & Strings of characters & \texttt{"Hello there"} \\
Symbol & Single-word strings, internally stored as integers and
generally used as hash keys & \texttt{`foo} \\
Code& Blocks of Angort code, anonymous functions & \texttt{( dup * 1 + )}\\
Integer range & 
A range of integers between two values with optional step &
\texttt{0 4 range}\footnotemark[1]\\
Float range & A range of floats between two values with 
step& \texttt{0 4 0.1 frange}\\
List & A array/list of values\footnotemark[2] & \texttt{[1,2,"foo"]} \\
Hash & A map of values to values implemented as a hash table, where
keys can strings, symbols, integers or floats & \texttt{[\% `foo 1, "fish" "fish"]} \\
\hline
\end{tabular}
\end{center}
\footnotetext[1]{This is actually a call to the \texttt{range} function which takes two integers to create a range.}
\footnotetext[2]{Stored internally as an
automatically resized array rather than a linked list.}
``Code'' above actually conflates two types internally --- codeblocks,
which have no environment; and closures, which consist of a codeblock and
some stored variables. They appear identical to the user.
There are some other types used internally, such as the
types for iterators and deleted objects in a hash.

Integers can also have a base character, `b', `x'/`h', `o' or `d', which
goes before the optional 'l' for 'long.' \textbf{Note that} all numbers
must start with a digit, so a leading zero will be required for some 
hex numbers (e.g. `0ffh' not `ffh').

\todo{Checking and getting types (isnone, isnumber, iscallable, type)}

\todo{heredocs}

\clearpage
\subsubsection{Coercions}
\begin{itemize}
\item Integers and floats are coerced to strings in string contexts.
\item In binary operations, if one of the operands is a float, both are coerced to floats:
\begin{v}
1|0> 1 2 / .
0
1|0> 1.0 2 / .
0.500000
\end{v}
\item In certain binary operations (currently just ``+'') if one of the operands
is a string, both will be coerced to strings.
\item \todo{more... longs and doubles, notably, and what 
happens when you can't}
\end{itemize}

\subsection{Defining new functions}
New functions are defined with code of the form
\begin{v}
:functionname ... ;
\end{v}
A simple example is:
\begin{v}
:square dup *;
\end{v}
\indw{dup}
This will define a function \texttt{square} which will duplicate
the value on top of the stack, multiply the top two values
(effectively squaring the number\footnote{This will produce
an error if the value is not a number.}) and exit, leaving the
squared number on the stack. This can then be used thus:
\begin{v}
3 square 4 square + .
\end{v}
which will print 25, i.e. $3^2+4^2$. Note that leaving values behind
on the stack on exiting a function is how we return values in Angort.
While defining a function, Angort is is compilation mode --- functions will
be converted to bytecode but added to the definition of the new function
rather than executed immediately. Function definitions can therefore span
more than one line:
\begin{lstlisting}
:factorial |x:|
    ?x 1 = if
        1
    else
        ?x ?x 1 - factorial *
    then
;  
\end{lstlisting}
This example uses recursion and the Angort \texttt{if..else..then} construction.
This looks a little odd to programmers not familiar with Forth -- please
see Section~\ref{conditions} for more details.
\todo{Mention how immediate and script mode work with definitions. Add parameter typing.}


\subsubsection{Word parameters and local variables}
Until now, all the code has been perfectly valid Forth\footnote{With
the exception of the factorial function, which uses local variables
and recursion.}. However, the manipulation of
stack required in Forth is a challenge (and modern computers
have a little more memory), so Angort has a system of named
parameters and local variables. These can be defined by
putting a special block of the form
\begin{v}
|param1,param2,param3... : local1,local2,local3|
\end{v}
after the function name in the definition. Locals and parameters are
exactly the same internally, but the values of parameters are popped
off the stack when the new function is called. 


Once defined, locals (and parameters) can be read (i.e. pushed onto
the stack) using a question mark followed by the variable name.
Similarly, a local is written (i.e. a value popped off the stack and
stored in the local) by using an exclamation mark. For example,
\begin{lstlisting}
:pointless |x,y:z|
    ?x ?y + !z
;
\end{lstlisting}
will read the two arguments into the locals $x$ and $y$, add them,
store the result in the local $z$, and then exit, throwing everything away.
Note that a function with parameters but no locals is defined by leaving
the part after the colon empty:
\begin{lstlisting}
:magnitude |x,y:|
    ?x dup *
    ?y dup * +
    sqrt
;
\end{lstlisting}
while leaving the part before the colon empty will define a function with
locals but no parameters:
\begin{lstlisting}
:countToTen |:count|
    0!count
    {
        ?count 1+ dup !count
        dup .
        10 = ifleave
    }
;
\end{lstlisting}
Here, the curly brackets enclose a loop, which is exited by the
\texttt{ifleave} instruction when the stack top value is 10. See
Section~\ref{loops} for more details.
\todo{Types in param lists. Returning values.}
     
\subsubsection{Word documentation strings and stack pictures}
\label{stackpic}
Following the locals and parameters (or just the function name if there
are none) there may be a function documentation string. It has the form
\begin{v}
:"(before -- after) what the function does"
\end{v}
The section in brackets is known as a \emph{stack picture,}\footnote{Or sometimes
a \emph{stackflow symbol}.} and describes
the action of the function on the stack. The part before the hyphen
shows the part of the stack removed by the function, and the part after
shows its replacement. For example:
\begin{lstlisting}
:dist |x1,y1,x2,y2:|
    :"(x1 y1 x2 y2 -- distance) calculate the distance between two points"
    ?x1 ?x2 - abs dup *     # this is (x1-x2)^2
    ?y1 ?y2 - abs dup *     # this is (y1-y2)^2
    + sqrt                  # sum them, and find the root
;
\end{lstlisting}
Note that the names in the picture are not necessarily those of variables:
they are single word descriptions of values. Typically these will be variable
names for the ``before'' part, but not for the ``after'' part (since
the return values are simply what's left on the stack).

To print the documentation for a word, enter \texttt{??wordname} at
the prompt. This will also work for most built-in Angort functions,
but not all\footnote{This is because most Angort functions compile
to calls to C++ functions, which can have descriptions. Simpler operations,
such as \texttt{if..then..else}, loop markers, arithmetic and the ``.'' print operator, compile directly to bytecode opcodes.
These cannot have descriptions.}. It will also work for any functions
inside C++ plugin libraries (see Section~\ref{library}). For example:
\begin{v}
Angort version 2.9.2 (c) Jim Finnis 2012-2017
Use '??word' to get help on a word.
1|0 > ??redir
redir: (filename --) open a new file and redirect output to it.
 For more complex file output, use the IO library. On fail, throws
 ex$failed.
1|0 > 
\end{v}


\section{Basic built-in words}
\subsection{Stack manipulation}
\index{stack!manipulation}
There exist a number of words whose sole purpose is to manipulate
the stack. These words, and their stack pictures, are given in the table
below.
\indw{dup}\indw{drop}\indw{swap}\indw{over}
\begin{center}
\begin{tabular}{|l|p{4in}|}\hline
\textbf{name} & \textbf{stack picture}\\ \hline
dup & (a -- a a)\\
drop & (a --)\\
swap & (a b -- b a)\\
over & (a b -- a b a)\\
\hline
\end{tabular}
\end{center}
Note that this is a much smaller set than the set of Forth stack
words, which includes \texttt{rot}, \texttt{roll}, \texttt{nip} and \texttt{tuck}.
The local variable system makes such complex operations unnecessary.

\subsection{Binary operators}
\indw{arithmetic}\indw{binary operations}
In the following operations, these conversions take place:
\begin{itemize}
\item If one of the operands is a string, the other will be converted to a string. Only ``+'', ``*''  and comparison operators will be valid.
\item If one of the operands is a long and the other is an integer, the integer will be converted to a long and the result will be a long.
\item If one of the operands is a float and the other is an integer or long, that will be converted to a float and the result will be a float.
\item If one of the operands is a double and the other is an integer, long, or float, that will will be converted to a double and the result will be a double.
\item The comparison operators will do identity checks on objects (lists, ranges etc.), not deep comparisons.
\item The comparison functions actually return integers, with non-zero indicating falsehood.
\end{itemize}
\indw{and}\indw{or}
\indw{band}\indw{bor}\indw{bxor}
\begin{center}
\begin{tabular}{|l|l|p{4in}|}\hline
\textbf{name} & \textbf{stack picture} & \textbf{side-effects and notes}\\ \hline
+    & (a b -- a+b)&\\
-    & (a b -- a-b)&\\
$*$    & (a b -- a*b)& ints are coerced to floats if one operand is an int, but not otherwise. ``string int *'' produces
a repeated string, but not ``int string *''.\\
/    & (a b -- a/b)&\\
\%    & (a b -- a\%b) & integer remainder ("mod") operator\\
$>$    & (a b -- a$>$b)& string comparison works as expected\\
$<$    & (a b -- a$<$b)& string comparison works as expected\\
=    & (a b -- a=b)& equality test: string comparison works as expected\\
!=   & (a b -- a!=b)& inequality test: string comparison works as expected\\
and & (a b -- a$\wedge$b) & binary and -- inputs are coerced to integers, nonzero is true\\
or & (a b -- a$\vee$b) & binary or  -- inputs are coerced to integers, nonzero is true\\
band & (a b -- a \& b) & bitwise AND \\
bor & (a b -- a $|$ b) & bitwise OR \\
bxor & (a b -- a $\sim$ b) & bitwise XOR \\
\hline
\end{tabular}
\end{center}

\subsection{Unary operators}
\indw{bnot}\indw{not}\indw{neg}\indw{abs}
\begin{center}
\begin{tabular}{|l|l|p{4in}|}\hline
\textbf{name} & \textbf{stack picture} & \textbf{side-effects and notes}\\ \hline
not & (a -- !a) & logical negation of a boolean\\
neg & (a -- -a) & arithmetic negation of float or integer\\
abs & (a -- $|\textrm{a}|$) & absolute value\\
isnone & (a -- bool) & true if value is \texttt{none} \\
bnot & (a -- !b) & bitwise not\\    
\hline
\end{tabular}
\end{center}

\subsubsection{Input and output}
\todo{Increase the meaning!}
\begin{itemize}
\item dot, p, nl
\item errp, errnl
\item redir, endredir
\item rawp, x
\item format
\end{itemize}

